---
title: "Exploratory data analysis of HELPrct data"
author: "YOUR NAME HERE"
date: "7/24/2020"
output: pdf_document
---


<!---
Steps for SSW 798 - Secondary Data Analysis:
1. Change your name above in the "Author" section where it says "YOUR NAME HERE" (keep the quotes)
2. Complete all the steps below that are included in comments that look like this and are in ALL CAPS (I'm not yelling, just trying to keep it clear...)
3. When you're done, click the "Knit" icon above, which will generate a pdf document with both text and statistical output
4. Open the PDF of the same name from the Files tab, ideally along side the screen you're looking at here.

[Note: this text will be hidden in the pdf document you generate. It's just between us!]
--->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mosaic)
data("HELPrct")
View(HELPrct)
```

## What is the value of `cesd` for the 5th Participant?

```{r}
head(HELPrct)
```

<!--- WRITE A COMPLETE SENTENCE BELOW THAT INTERPRETS THE VALUE OF CESD FOR THE 5TH OBSERVATION / PARTICIPANT. A VALUE OF 20 FOR A CESD CAN BE INTERPRETED AS A SCORE OF 20. --->


### Based on what you now know about the CESD measure, do you think Participant #5's score is high or low?

<!--- WRITE A COMPLETE SENTENCE BELOW THAT ANSWERS THE QUESTION ABOVE --->


## How (a)typical is the 5th Participant's CESD score in this population?

<!--- RUN THE CODE CHUNK BELOW AND USE THE OUTPUT TO ANSWER THE QUESTION. --->
```{r pressure, echo=FALSE, message=FALSE}
library(tidyverse)
qplot(x=cesd, data=HELPrct)
```

The y-axis is the number of people, the x-axis is the CESD score, and the height of the bar reflects how many people in this study had that CESD score.


## Based on the graph above, do you think Participant #5's score is high or low? Why or why not?

<!--- WRITE A COMPLETE SENTENCE BELOW THAT INTERPRETS THE VALUE OF CESD FOR THE 5TH OBSERVATION / PARTICIPANT RELATIVE TO THE DISTRIBUTION OF ALL CESD SCORES IN THE SAMPLE AND EXPLAINS YOUR RATIONALE--->



<!--- FINAL STEP! CHANGE YOUR NAME IN THE VERY TOP SECTION OF THE DOCUMENT AND THEN CLICK THE "KNIT" ICON JUST ABOVE THE TOP OF THE DOCUMENT TO CONVERT THIS INTO A PDF. --->
